Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-11-12 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@665 3fc257bb-8550-0410-81b3-a5c103da83f7

The algorithm in Figure~\ref{fig:pr2} performs these iterations, using
linear-algebra concepts implemented in a MapReduce framework, with
data stored in 4 MapReduce objects: $A$, $MT$, $x$, and $y$.  $A$
stores the matrix, $MT$ its ``empty'' rows, while $x$ and $y$ are
vectors of length $|V|$.  The nonzeros of matrix $A$ are initialized
as described above; the PageRank vector $x$ is initialized uniformly.
A MapReduce object $MT$ contains the indices of all-zero rows of $A$,
corresponding to vertices with no outlinks.  The PageRank algorithm is
made more efficient by invoking the {\it aggregate()} operation once
on the MapReduce objects representing the matrix $A$, the all-zero
matrix rows $MT$, and the vector $x$, before the PageRank iterations
begin.  Since these 3 MapReduce objects are all keyed by vertices,
pre-aggregating them moves all KV pairs with a given key to the same
processor.  Many of the subsequent PageRank operations then become
local operations, replacing {\it collate()} (which is equivalent to an
{\it aggregate()} followed by {\it convert()}) with simply a {\it
convert()}.  In Figure~\ref{fig:pr2}, steps where {\it convert()} is
substituted for {\it collate()} are marked with an asterisk.  This
strategy of pre-aggregating to improve data locality is useful in many
MapReduce algorithms, but assumes that identical keys in different
MapReduce objects can be mapped to the same processor.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute contribution for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $(x_i, NULL)$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$ \\
\> \> \> $c_{local} = c_{local} + foo$ \\
\> \> \> Output:  If ${nvalues} = 2$: Key = $i$, Value = NULL \\
\> \> $c_{global}$ = MPI\_Allreduce($c_{local}$,MPI\_SUM) \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, Multivalue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, Multivalue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{global}$ to $y$; compute $n_{global}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{global}$; $n_{local} = {\rm max}(y_i,n_{local})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $n_{global}$ = MPI\_Allreduce($n_{local}$,MPI\_SUM) \\
\> 4 Scale $y$ by $n_{global}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, Multivalue = $x_i, y_i$ \\
\> \> \> $r_{local} = {\rm max}(|x_i - y_i|,r_{local})$ \\
\> \> $residual$ = MPI\_Allreduce($r_{local}$,MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank \\

@@

In Step 1 of a PageRank interation, $MT$ is used in dot products to
compute adjustments to the PageRank vector to represent the uniform
probability of going from a leaf page to any other page.  Adjustments
for random jumps from any page are also computed.  In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual to determine whether the PageRank vector has converged;
work is again proportional to $|V|$.  In Step 6, the PageRank vector
is overwritten prior to the next iteration.  Throughout the algorithm,
we exploit the ability to call MPI\_Allreduce to compute global norms
and residuals.

@@

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration and
vertex distancs are updated.  A MapReduce formulation of this
algorithm, described in Figure~\ref{fig:sssp}, stores directed edges
in $E$ and begins by labeling the distance to all vertices as infinte,
except a distance of 0 to the source vertex $V_s$.  At each iteration,
a {\it map()} is performed using $E$ as input to add edges and their
weights to each vertex in $V$.  After a {\it collate()}, each vertex
$V_i$ knows its outgoing edges and their weights, as well as its own
current distance.  It may also have one or more distances assigned to
it from a previous iteration (from an inbound edge).  In the {\it
reduce()}, each vertex resets its distance to the minimum of these new
distances and its current distance.  If its distance changed, it emits
new distance values $d_j$ for the vertices associated with its
outgoing edges, where $d_j = d_i + w_{ij}$.  It also unsets a ``done''
flag, stored by each processor.  When the reduce is complete, an
MPI\_Allreduce is performed to check the ``done'' flag status on all
processors.  If any distance was changed, the iterations continue, and
the distance updates propagate through the graph.  If no distances
changed, the algorithm is done.

>>>>

The algorithm in Figure~\ref{fig:pr2} performs these iterations, using
linear-algebra concepts implemented in a MapReduce framework, with
data stored in 4 MapReduce objects: $A$, $MT$, $x$, and $y$.  $A$
stores the matrix, $MT$ its ``empty'' rows, while $x$ and $y$ are
vectors of length $|V|$.  The nonzeros of matrix $A$ are initialized
as described above; the PageRank vector $x$ is initialized uniformly.
A MapReduce object $MT$ contains the indices of all-zero rows of $A$,
corresponding to vertices with no outlinks.  The PageRank algorithm is
made more efficient by invoking the {\it aggregate()} operation once
on each of these MapReduce objects before the PageRank iterations
begin.  Since the 3 objects are all keyed by vertex, pre-aggregating
them moves all KV pairs with a given key to the same processor.  Many
of the subsequent PageRank operations then become local operations,
replacing a {\it collate()} (which is equivalent to an {\it
aggregate()} followed by {\it convert()}) with simply a {\it
convert()}.  In Figure~\ref{fig:pr2}, steps where {\it convert()} is
substituted for {\it collate()} are marked with an asterisk.  This
strategy of pre-aggregating to improve data locality is useful in many
MapReduce algorithms, but assumes that identical keys in different
MapReduce objects can be mapped to the same processor.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute contribution for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$, MultiValue = $(x_i, NULL)$ if $i \in {MT}$ else MultiValue = $x_i$ \\
\> \> \> If ${nvalues} = 2$: $c_{dot} = c_{dot} + x_i$ \\
\> \> \> Output: Key = $i$, Value = NULL if ${nvalues} = 2$ \\
\> \> $c_{dot}$ = MPI\_Allreduce($c_{dot}$,MPI\_SUM) \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, Multivalue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, Multivalue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{dot}$ to $y$; compute $y_{max}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{dot}$; $y_{max} = {\rm max}(y_i,y_{max})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $y_{max}$ = MPI\_Allreduce($y_{max}$,MPI\_MAX) \\
\> 4 Scale $y$ by $y_{max}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / y_{max}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, Multivalue = $x_i, y_i$ \\
\> \> \> $residual = {\rm max}(|x_i - y_i|,residual)$ \\
\> \> $residual$ = MPI\_Allreduce($residual$,MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank \\

@@

In Step 1 of a PageRank interation, $MT$ is effectively dotted with
$x$ to compute an adjustment $c_{dot}$ to the PageRank vector that
represents the uniform probability of going from a leaf page to any
other page.  $MT$ is unchanged by the operation.  Adjustments for
random jumps from any page are also computed.  In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual as an infinity norm to determine whether the PageRank
vector has converged; work is again proportional to $|V|$.  Note that
$y$ and $x$ are both keyed by vertex at this point, so they can be
joined via an {\it add()} and {\it convert()} without need for the
communication required by a full {\it collate()}.  In Step 6, the
PageRank vector is overwritten prior to the next iteration.
Throughout the algorithm, we exploit the ability to call
MPI\_Allreduce to compute global dot products and norms.

@@

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration and
vertex distancs are updated.  A MapReduce formulation of this
algorithm, described in Figure~\ref{fig:sssp}, begins with a graph $E$
of directed edges $(V_i,V_j)$.  As generated by the R-MAT procedure
disussed earlier, this may include self-edges $(V_i,V_i)$ and loop
edges (both $(V_i,V_j)$ and $(V_j,V_i)$).  These are handled as
special cases within the algorithm to avoid spurious contributions to
the shortest path distances.  MapReduce object $V$ stores the distance
of each vertex $V_i$ from the source vertex $V_s$.

To begin, the distance to all vertices is set to infinity, except a
distance of 0 for $V_s$ itself.  At each iteration, a {\it map()} is
performed using $E$ as input to add edges and their weights to each
vertex in $V$.  After a {\it collate()}, each vertex $V_i$ knows its
outgoing edges and their weights, as well as its own current distance.
It may also have one or more distances assigned to it from a previous
iteration (from an inbound edge).  In the {\it reduce()}, each vertex
resets its distance to the minimum of these new distances and its
current distance.  If its distance changed, it emits new distance
values $d_j$ for the vertices associated with its outgoing edges,
where $d_j = d_i + w_{ij}$.  It also unsets a ``done'' flag, stored by
each processor.  When the {\it reduce()} is complete, an
MPI\_Allreduce is performed to check the ``done'' flag status on all
processors.  If any distance was changed, the iterations continue, and
the distance updates propagate through the graph.  Otherwise the
algorithm is done.

### 2010-11-12 paper/lessons.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@665 3fc257bb-8550-0410-81b3-a5c103da83f7

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a data
shuffle or {\it collate()} operation) with communication between all
pairs of processors.  But the benefit is often good load-balance, even
for hard-to-balance irregular data sets, such as those derived from
sparse graphs with power-law degree distributions.  By contrast, more
traditional distributed-memory parallel algorithms, e.g. for matrix
operations, or grid-based or particle-based simulation codes,
typically work hard to localize data and minimize communication.  To
do this they often require application-specific logic and parallel
communication coding to create and maintain a data decomposition,
generate ghost copies of nearby spatially-decomposed data, etc.

@@

Finally, though this paper focuses on graph algorithms expressed as
MapReduce operations, there is nothing about the MR-MPI library itself
that is graph specific.  We hope the library can be generally useful
on large-scale monolithic or cloud-style parallel machines that
support MPI, for a variety of data-intensive or compute-intensive
problems that are amenable to solution using a MapReduce paradigm.

>>>>

MapReduce achieves parallelism through randomizing the distribution of
data across processors, which often intentionally ignores data
locality.  This translates into maximal data movement (during a data
shuffle or {\it collate()} operation) with communication between all
pairs of processors.  But the benefit is often good load-balance, even
for hard-to-balance irregular data sets, such as those derived from
sparse graphs with power-law degree distributions.  By contrast,
MPI-based parallel algorithms, e.g. for matrix operations, or
grid-based or particle-based simulation codes, typically work hard to
localize data and minimize communication.  To do this they often
require application-specific logic and parallel communication coding
to create and maintain a data decomposition, generate ghost copies of
nearby spatially-decomposed data, etc.

@@

Finally, though this paper focuses on graph algorithms expressed as
MapReduce operations, there is nothing about the MR-MPI library itself
that is graph specific.  We hope the library can be generally useful
on large-scale monolithic or cloud-style parallel machines that
support MPI, for a variety of data-intensive or compute-intensive
problems that are amenable to solution using a MapReduce approach.
